---
title: "Model building and Simulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
theme_set(theme_minimal())
```

## Model Building: DLMs

DLMs are represented as a `case class` containing the observation matrix and the system matrix:

```tut:silent
import core.dlm.model._
import breeze.linalg.{DenseMatrix, DenseVector}

case class Model(
  f: Double => DenseMatrix[Double],
  g: Double => DenseMatrix[Double]
)
```

The object `Dlm` contains functions to build several common models:

```tut:silent
val linear = Dlm.polynomial(1)
val seasonal = Dlm.seasonal(period = 24, harmonics = 3)
```

```scala
val x: Array[Dlm.Data] = ??? // timeseries to regress against
val regression = Dlm.regression(x = x)
```

These simple models can be composed to produce more complex models, for instance a seasonal model with a linear trend can be produced:

```tut:silent
import cats.implicits._

val seasonalTrend = linear |+| seasonal
```

The models can also be combined using a cartesian product in order to model multivariate time series:

```tut:silent
// model for a bivariate time series
val multivariateModel = linear |*| seasonalTrend

// model for 30 seemingly unrelated time series
val sutse = List.fill(30)(linear).reduce(_ |*| _)
```

## Model Building: DGLMs

A dynamic generalised linear model is a DLM with a non-Gaussian observation distribution. The observation distribution is from the Exponential family, although extensions can be made to incorporate the Student's t-distribution from the location-scale family.

To build a DGLM, first build a univariate DLM as above then select an observation model from the `Dglm` object:

```tut:silent
val poisson = Dglm.poisson(linear)
```

## Simulating from a DLM

Simulation is useful to test inference algorithms or to get a feel for how a model behaves with different sets of parameters. First we must define the model `Parameters`:

```tut:silent
val p = Dlm.Parameters(
  v =  DenseMatrix(1.0),
  w =  DenseMatrix(3.0),
  m0 = DenseVector(0.0),
  c0 = DenseMatrix(10.0)
)
```

Then we can simulate from the model:

```tut:silent
val data = Dlm.simulateRegular(0, linear, p, 1.0).
  steps.
  take(1000)
```

```{r first-order-simulated}
data = read_csv("../examples/data/first_order_dlm.csv")

data %>%
  gather(key, value, -time) %>%
  ggplot(aes(x = time, y = value, colour = key)) +
  geom_line() +
  theme(legend.position = "bottom")
```

Simulation from any DLM is possible using the same syntax. The parameters must be appropriate for the model, for instance the latent-state of a seasonal model has dimension of twice the number of harmonics in the model, hence `m0`, `c0` and `w` must be of the same dimension. 

## Simulating from a DGLM

In order to simulate from a DGLM we must know how to sample from the chosen observation distribution of the DGLM. The `Dglm.Model` class and its associated model building functions provide this function and an associated `simulate` function. The figure below shows a simulation from the Poisson model.

```tut:silent
Dglm.simulate(poisson, p)
```

```{r poisson-sims, fig.cap="Simulation from the poisson DGLM with $(W, m_0, C_0) = (0.01, 0.0, 1.0)$"}
data = read_csv("../examples/data/poisson_dglm.csv")

data %<>%
  gather(key, value, -time) %>%
  filter(time > 900) 

ggplot() +
  geom_line(data = subset(data, key == "state"), aes(x = time, y = value)) +
  geom_step(data = subset(data, key == "observation"), aes(x = time, y = value)) +
  facet_wrap(~key, scales = "free_y", ncol = 1) +
  theme(legend.position = "none")
```

Note that the Poisson distribution has only one parameter, the rate and hence `v` in the `Dlm.Parameters` specification is unused. For observation distributions with additional parameters, then the `v` is used for this purpose.
