---
title: "Online Parameter Inference"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(coda)
library(ggmcmc)
library(patchwork)
library(latex2exp)
theme_set(theme_minimal())
source("mcmc_summarise.R")
```

Given a DLM, with observations $Y_{1:N}$ and latent state $x_{0:N}$ with observation and state equations:

\begin{align*}
Y_t &= F_t^Tx_t + v_t, & v_t &\sim \mathcal{N}(0, V_t), \\
X_t &= G_tx_{t-1} + w_t, & w_t &\sim \mathcal{N}(0, W_t), \\
x_0 &\sim \mathcal{N}(m_0, C_0)
\end{align*}

We can joint posterior distribution of the parameters and latent-state, $p(\psi_t, x_t|y_{1:N})$ using on of the following online algorithms. These are useful for streaming data as they don't require re-running of costly MCMC algorithms as each new data point is observed. 

The online filters will be demonstrated for a first order polynomial DLM, with $V = 2, W = 3$, plotted in the figure below:

```{r first-order-simulated-online}
first_order = read_csv("../examples/data/first_order_dlm.csv")

first_order %>%
  gather(key, value, -time) %>%
  ggplot(aes(x = time, y = value, colour = key)) +
  geom_line() +
  theme(legend.position = "bottom") +
    labs(title = "Simulated values from the first order DLM",
         subtitle = TeX("$$W = 3$, V = 2$"))
```

## Conjugate Filter

The Conjugate filter is an exact filter, which augments the Kalman Filter to learn the observation variance using a conjugate update of the Gamma distribution.

```tut
val prior = InverseGamma(3.0, 4.0)
val filtered = ConjugateFilter(prior, ConjugateFilter.advanceState(p, mod.g))
  .filter(mod, data, p)
```

```{r}
conjugate_filter = read_csv("../examples/data/first_order_dlm_conjugate_filtered.csv")

p1 = conjugate_filter %>%
    mutate(upper = v_mean + 2 * sqrt(v_variance),
         lower = v_mean - 2 * sqrt(v_variance)) %>%
    ggplot(aes(x = time, y = v_mean)) +
    geom_line() +
    geom_hline(aes(yintercept = 2.0), linetype = "dashed") +
    geom_ribbon(aes(x = time, ymin = lower, ymax = upper), alpha = 0.5) +
    ylab("Observation Variance")

p2 = conjugate_filter %>%
    inner_join(first_order, by = "time") %>%
    mutate(upper = qnorm(p = 0.995, mean = state_mean, sd = sqrt(state_variance)),
           lower = qnorm(p = 0.005, mean = state_mean, sd = sqrt(state_variance))) %>%
    gather(key, value, state, state_mean) %>%
    ggplot(aes(x = time, y = value, colour = key)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower, ymax = upper), colour = NA, alpha = 0.5)

p1 / p2
```

## Liu and West Filter

The Liu and West filter can be used for general state space models:

```tut
val n = 500
  // smoothing parameter for the mixture of gaussians, equal to (3 delta - 1) / 2 delta
  val a = (3 * 0.95 - 1) / 2 * 0.95

  val prior = for {
    v <- InverseGamma(3.0, 4.0)
    w <- InverseGamma(3.0, 10.0)
  } yield DlmParameters(DenseMatrix(v), DenseMatrix(w), p.m0, p.c0)

  val filtered = LiuAndWestFilter(n, prior, a, n0).filter(mod, data, p)
```

```{r liu-and-west}
true_values = data_frame(param = c("v", "w"), value = c(2.0, 3.0))

liu_and_west = read_csv("~/git/bayesian_dlms/examples/data/liuandwest_filtered.csv")

liu_and_west %>%
    select(time, v = v_mean, w = w_mean) %>%
    gather(key = param, value = mean, v, w) %>%
    inner_join(liu_and_west %>%
    select(time, v = v_variance, w = w_variance) %>%
    gather(key = param, value = variance, v, w)) %>%
    rowwise() %>%
    mutate(upper = mean + 2 * sqrt(variance), lower = mean - 2 * sqrt(variance)) %>%
    select(time, param, mean, upper, lower) %>%
    ggplot(aes(x = time, y = mean)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) +
    geom_line() +
    geom_hline(data = true_values, aes(yintercept = value), 
               linetype = "dashed", colour = "#666666") +
    facet_wrap(~param, scales = "free_y", ncol = 1)
```

## RB Filter

The Rao-Blackwell filter marginalises the Gaussian part of the DLM in order to perform filtering using the Kalman filter conditional on sampled values of the static parameters. Parameter distributions at each time point are approximated by a weighted particle cloud. A Kalman filter is performed for conditional on the value of each particle. Thereby determining the join posterior of the parameters and latent state.

```tut
// smoothing parameter for the mixture of gaussians
val delta = 0.99
val a = (3 * delta - 1) / 2 * delta

val prior = for {
  v <- InverseGamma(3.0, 4.0)
  w <- InverseGamma(3.0, 10.0)
} yield DlmParameters(DenseMatrix(v), DenseMatrix(w), p.m0, p.c0)

val filtered = RaoBlackwellFilter(500, prior, a, 250).filter(mod, data, p)
```

```{r rb-filter}
rb_filter = read_csv("~/git/bayesian_dlms/examples/data/fo_raoblackwellfilter.csv")

# RB Parameters
p1 = rb_filter %>%
    select(time, v = v_mean, w = w_mean, contains("variance")) %>%
    gather(key = param, value = mean, v, w) %>%
    inner_join(rb_filter %>%
    select(time, v = v_variance, w = w_variance) %>%
    gather(key = param, value = variance, v, w)) %>%
    rowwise() %>%
    mutate(upper = mean + 2 * sqrt(variance), lower = mean - 2 * sqrt(variance)) %>%
    ggplot(aes(x = time, y = mean)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) +
    geom_hline(data = true_values, aes(yintercept = value), linetype = "dashed", colour = "#666666") +
    facet_wrap(~param, scales = "free_y", ncol = 1, strip.position = "right")

# RB State
p2 = rb_filter %>%
    inner_join(first_order, by = "time") %>%
    gather(key, value, contains("state")) %>%
    ggplot(aes(x = time, y = value, colour = key)) +
    geom_line(alpha = 0.75) +
    theme(legend.position = c(0.9, 0.2))

p1 / p2
```

## Storvik Filter

Using conjugate updating for $V$ and $W$:

```tut
val priorV = InverseGamma(3.0, 4.0)
val priorW = InverseGamma(3.0, 10.0)

val filtered = StorvikFilter.filterTs(mod, data, p, priorV, priorW, n, n)
```

```{r storvik-filter}
storvik = read_csv("~/git/bayesian_dlms/examples/data/fo_storvik_filtered.csv")

storvik %>%
    select(time, v = v_mean, w = w_mean) %>%
    gather(key = param, value = mean, v, w) %>%
    inner_join(storvik %>%
    select(time, v = v_variance, w = w_variance) %>%
    gather(key = param, value = variance, v, w)) %>%
    rowwise() %>%
    mutate(upper = mean + 2 * sqrt(variance), lower = mean - 2 * sqrt(variance)) %>%
    ggplot(aes(x = time, y = mean)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) +
    geom_hline(data = true_values, aes(yintercept = value), linetype = "dashed", colour = "#666666") +
    facet_wrap(~param, scales = "free_y", ncol = 1)

```

