---
title: "Particle Filter"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(ggmcmc)
library(coda)
library(latex2exp)
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
theme_set(theme_minimal())
```

## Particle Filter

The particle filter can be used for non-linear and non-Gaussian state space models, such as DGLMs. A general DGLM is given by:

$$\begin{align*}
Y_t &\sim p(y_t | g(F_t^Tx_t), \theta) \\
X_t &= G_tX_{t-1} + w_t, \qquad w_t \sim \mathcal{N}(0, W), \\
X_0 &\sim \mathcal{N}(m_0, C_0).
\end{align*}$$

Where $p(y_t | F_t^Tx_t, \theta)$ is an Exponential Family distribution, with optional static parameter $\theta$.

The particle filter can be used to find the filtering state $p(x_{0:t}|y_{1:T}, \psi)$, where $\psi$ represents the static parameters in the model. The particle filter algorithm proceeds as follows:

*Initialisation*
1. Sample $k$ initial particles from the initial distribution, $x_0^{(i)} \sim \mathcal{N}(m_0, C_0), i = 1,\dots,k$
2. Initialise the particle weights $w^{(i)} = \frac{1}{k}$

*Main Loop*

3. Assume at time $t$ we have weighted sample from the posterior $\{x_t^{(i)}, w_t^{(i)}, i = 1,\dots,k\}$ 
4. Advance the state using the state evolution kernel $x_{t+1}^{(i)} \sim \mathcal{N}(G_tx_t^{(i)}, W), i = 1,\dots,k$
5. Calculate the weights using the conditional likelihood: $w_{t+1}^{(i)} = p(y_{t+1}|x_{t+1}^{(i)}), i = 1,\dots,k$
7. Normalise the weights, $\pi_{t+1}^{(i)} = \frac{w_{t+1}^{(i)}}{\sum_{j=1}^kw_{t+1}^{(j)}}, i = 1,\dots,k$
6. Resample the particles by sampling $u_j \sim p(\pmb\pi_{t+1})$ then select $x^{(u_j)}_{t+1}$ $j = 1,\dots,k$

Repeat the mean loop until $t = T$.

To apply the particle filter to a model using this library, specify either a DGLM or a DLM model. Consider a poisson model

```tut
import com.github.jonnylaw.dlm._
import breeze.linalg.{DenseMatrix, DenseVector}
val mod = Dglm.poisson(Dlm.polynomial(1))
val params = DlmParameters(DenseMatrix(2.0),
                           DenseMatrix(0.05),
                           DenseVector(0.0),
                           DenseMatrix(1.0))

val sims = Dglm.simulateRegular(mod, params, 1.0).
             steps.take(100).map(_._1).toVector
```

A simulation from the Poisson model is presented in the figure below:

```{r poisson-dglm}
poisson_sims = read_csv("../examples/data/poisson_dglm.csv")

data = poisson_sims %>%
  gather(key, value, -time)

ggplot() +
  geom_line(data = subset(data, key == "state"), aes(x = time, y = value)) +
  geom_step(data = subset(data, key == "observation"), aes(x = time, y = value)) +
  facet_wrap(~key, scales = "free_y", ncol = 1) +
  theme(legend.position = "none")
```

Then the bootstrap particle filter can be user to determine the latent-state of the partially observed inhomogeneous Poisson process:

```tut
val n = 500
val filtered = ParticleFilter(n, n, ParticleFilter.metropolisResampling(10)).filter(mod, sims, params)
```

```{r}
filtered = read_csv("../examples/data/poisson_filtered_metropolis.csv")

p1 = poisson_sims %>%
    ggplot(aes(x = time, y = observation)) +
    geom_step()

p2 = poisson_sims %>%
    inner_join(filtered) %>%
    mutate(upper = state_mean + 2 * sqrt(state_var),
           lower = state_mean - 2 * sqrt(state_var)) %>%
    gather(key, value, state, state_mean) %>%
    ggplot(aes(time, value, colour = key)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5, colour = NA)

p1 / p2
```
