---
title: "Investment Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(coda)
library(ggmcmc)
theme_set(theme_bw())
```

## R Markdown

This example is reproduced from [DLMs with R](https://www.springer.com/us/book/9780387772370), Section 4.5.1.

```{r read_investment_data}
data = read_csv("../data/invest2.dat", col_names = c("Denmark", "Spain"))

data %>%
  mutate(time = 1960:(1960 + nrow(data) - 1)) %>%
  gather(key = Country, value, -time) %>%
  ggplot(aes(x = time, y = value)) +
  geom_line() +
  facet_wrap(~Country, scales = "free_y", ncol = 1)
```

# Proposed Model

The proposed model is the second order DLM representing a linear growth.

$$\begin{align}
Y_t &= F \textbf{x}_t + \textbf{v}_t, \quad v_t \sim \mathcal{N}(0, V), \\
\textbf{X}_t &= G \textbf{x}_{t-1} + w_t, \quad w_t \sim \textrm{MVN}(0, W), \\
\textbf{X}_0 &\sim \textrm{MVN}(m_0, C_0).
\end{align}$$

The state is two dimensional, as such the system noise matrix $W$ is a $2 \times 2$ matrix. The observation and system evolution matrices do not depend on time, the observation matrix is $F = (1 \quad 0)$ and the system evolution matrix is:

$$G = \begin{pmatrix} 
1 & 1 \\
0 & 1
\end{pmatrix}.$$

# Gibbs Sampling

First we consider the joint distribution of the random variables in the model:

$$p(\textbf{x}_{0:T}, y_{1:T}, V, W) = p(V)p(W)\prod_{t=1}^T p(y_t|\textbf{x}_t, V)p(\textbf{x}_t|\textbf{x}_{t-1}, W).$$

The joint distribution simplifies nicely because of the conditional independent of the observations given the latent-state and the Markov property of the latent-state. In order to construct a gibbs sampler for the process, we can sample the latent-state using forward filtering backward sampling and use Inverse Gamma distributions for the prior distributions of $V$ and $W$. This reduces to a conditional conjugate update for $W$ of the form:

$$p(W|\textbf{x}_{0:T}, y_{1:T}, V) = \textrm{Inv-Gamma}\left(\alpha + \frac{T}{2}, \beta + \frac{1}{2}\sum_{t=1}^T (\textbf{x}_t - G_t\textbf{x}_{t-1})\right)$$

And for the observation matrix, $V$:

$$p(V|\textbf{x}_{0:T}, y_{1:T}, W) = \textrm{Inv-Gamma}\left(\alpha + \frac{T}{2}, \beta + \frac{1}{2}\sum_{t=1}^T (y_t - F_t\textbf{x}_{t})\right)$$

It is useful to specify prior belief about the parameters in terms of the mean and variance. If $X \sim \textrm{Inv-Gamma}(\alpha, \beta)$ then:

$$\begin{aligned}
\mathbb{E}(X) &= \frac{\beta}{\alpha - 1}, \\
\textrm{Var}(X) &= \frac{\beta^2}{(\alpha-1)^2(\alpha-2)}.
\end{aligned}$$

Then let $\mathbb{E}(X) = a$ and $\textrm{Var}(X) = b$ and the solve for $\alpha$ and $\beta$:

$$\begin{aligned}
\alpha &= \frac{2b + a^2}{b}, \\
\beta &= \frac{a(a^2+b)}{b}.
\end{aligned}$$

The prior mean for the observation variance is $a_y = 1$ with variance $b_y = 1000$ and the prior mean for the system variance is $a_x = 0.1$ and variance $b_x = 1000$ leading to prior distributions of:

```{r calculate_alpha_beta}
alpha = function(a, b) {
  (2 * b + a ** 2 ) / b
}

beta = function(a, b) {
  (a/b) * (a **2 + b)
}
```


$$\begin{aligned}
p(V) &= \textrm{Inv-Gamma}(2.001, 1.001) \\
p(W) &= \textrm{Inv-Gamma}(2.00001, 0.100001)
\end{aligned}$$

The following figure shows the density of these prior distributions:

```{r plot_priors}
prior_v = data_frame(x  = 1 / rgamma(n = 1000, shape = alpha(1, 1000), rate = beta(1, 1000)))
prior_w = data_frame(x = 1 / rgamma(n = 1000, shape = alpha(0.1, 1000), rate = beta(0.1, 1000)))

p1 = ggplot(prior_v, aes(x = x)) + 
  geom_density() +
  coord_cartesian(xlim = c(0, 3)) +
  labs(title = "Density of the prior distribution of V")

p2 = ggplot(prior_w, aes(x = x)) + 
  geom_density() +
  coord_cartesian(xlim = c(0, 1)) +
  labs(title = "Density of the prior distribution of W")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

The posterior densities summarised from 12,000 MCMC draws are plotted below:

```{r read_iters}
gibbs_iters = read_csv("../data/gibbs_spain_investment.csv")
```


```{r plot_gibbs_distributions}
params = gibbs_iters %>% mcmc() %>% ggs()

p1 = ggs_traceplot(params) + facet_wrap(~Parameter, strip.position = "top", ncol = 3)

p2 = ggs_histogram(params) +
  facet_wrap(~Parameter, ncol = 3) +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    legend.position = "none"
  )

p3 = ggs_autocorrelation(params) + 
  facet_wrap(~Parameter, ncol = 3) + 
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
  )

gridExtra::grid.arrange(p1, p2, p3, ncol = 1)
```