---
title: "Stochastic Volatility"
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(ggmcmc)
library(coda)
library(latex2exp)
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
theme_set(theme_minimal())
source("mcmc_summarise.R")
```

Stochastic volatility models are a class of models for non-stationary time series data. The log of the variance of the series is assumed to evolve according to a latent Gaussian process. The equation below shows a univariate stochastic volatility model with an AR(1) latent state:

\begin{align*}
Y_t &= \varepsilon_t\exp(\alpha_t/2), &\varepsilon_t &\sim \mathcal{N}(0, 1), \\
\alpha_t &= \mu + \phi(\alpha_{t-1} - \mu) + \eta_t, &\eta_t &\sim \mathcal{N}(0, \sigma_\eta^2).\numberthis \label{eq:usv-spec}
\end{align*}

$Y_t$, $t = 1,\dots,T$ are discrete observations of a time series process. The latent state represents the log-volatility and $\alpha_t$ follows an AR(1) process with mean $\mu$.

The figure below shows a simulation from this model with fixed
values of the parameters, $\phi = 0.8, \mu = 1, \sigma_\eta = 0.1$, the initial
state of the log-volatility is the stationary solution of the AR(1) process:
$\alpha_0 \sim \mathcal{N}\left(\mu, \frac{\sigma_\eta^2}{1 - \phi^2}\right)$.

To define this model in Scala and simulate observations, first import the required libraries and use the method simulate which initialises a Breeze `MarkovChain` object with the initial state equivalent to the stationary solution of the AR(1) process. 
 
```tut
import dlm.core.model._
val p = SvParameters(phi = 0.8, mu = 2.0, sigmaEta = 0.3)
val sims = StochasticVolatility.simulate(p).steps.take(1000).toVector.tail
```

```{r}
sims = read_csv("~/git/bayesian_dlms/examples/data/sv_sims.csv")

sims %>%
  filter(time < 1000) %>%
  gather(key, value, -time) %>%
  ggplot(aes(x = time, y = value)) +
  geom_line() +
  facet_wrap(~key, ncol = 1, scales = "free_y") +
  labs(title = TeX(string = "$\\phi = 0.8$, $\\mu = 2$, $\\sigma_{\\eta} = 0.3$"))
```


## Parameter Inference

To determine the posterior distribution of the parameters specify the prior distributions of the parameters:

```tut
val priorPhi = Gaussian(0.8, 0.1)
val priorMu = Gaussian(2.0, 1.0)
val priorSigma = InverseGamma(2.0, 2.0)
```

There are two inference algorithms defined for the stochastic volatility model, either one can be used and should produce identical inferences:

1. Likelihood Analysis of Non-Gaussian Measurement Time Series: http://www.jstor.org/stable/2337586
2. Stochastic volatility: likelihood inference and comparison with ARCH models: https://faculty.mccombs.utexas.edu/carlos.carvalho/teaching/KSB98-SV.pdf

```tut
val iters1 = StochasticVolatilityKnots.sampleParametersAr(priorPhi,
                                                           priorMu,
                                                           priorSigma,
                                                           sims)
val iters2 = StochasticVolatility.sampleUni(priorPhi, priorMu, priorSigma, sims)
```

The figure below shows the diagnostic plots for the posterior distribution of the parameters of the univariate stochastic volatility model.

```{r stochastic-volatility-parameters}
actual_values = tibble(
  Parameter = c("phi", "mu", "sigma"),
  actual_value = c(0.8, 2.0, 0.3)
)

files = c("../examples/data/sv_params_0.csv", 
          "../examples/data/sv_params_1.csv")
chains = read_chains(files, param_names = actual_values$Parameter, nth = 20, drop = 1e4)

plot_diagnostics_sim(chains, actual_values)
```